---
title: "Machine Learning: Mathematical Theory and Applications"
subtitle: ""
author: 
  - Sebastian Galeano
date: last-modified
format: 
  html:
    self-contained: true
toc: true
execute:
  error: false
theme: Default
title-block-banner-color: Primary
editor: visual
---

```{=html}
<style>
.boxed-text {
  border: 2px solid black;
  padding: 10px;
  margin: 10px 0;
}
</style>
```
```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## 1. Linear discriminant analysis for cancer data

#### 👾 Problem 1.1

::: boxed-text
Derive (analytically) $p(\mathbf{x})$ for the example above.
:::

1.  **The Joint Probability of** $\mathbf{x}$ **and Class** $y$

Given that $\mathbf{x} | y = m \sim \mathcal{N}(\boldsymbol{\mu}_m, \boldsymbol{\Sigma})$, which implies that the conditional probability density function for each class is:

$$p(\mathbf{x} | y = m) = \frac{1}{(2\pi)^{d/2} |\boldsymbol{\Sigma}|^{1/2}} \exp\left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_m)^\top \boldsymbol{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu}_m) \right)$$

where $d = 2$ for the two features "radius" and "texture."

2.  **The Marginal Density** $p(\mathbf{x})$

We need to marginalize over each class to obtain $p(\mathbf{x})$. We can write $p(\mathbf{x})$ as:

$$p(\mathbf{x}) = \sum_{m=1}^{2} p(\mathbf{x} | y = m) \pi_m$$

where $\pi_m = \Pr(y = m)$ are the class membership probabilities.

Thus, $p(\mathbf{x})$ is the weighted sum of the two Gaussian distributions corresponding to the two classes. Specifically:

$$p(\mathbf{x}) = \pi_1 p(\mathbf{x} | y = 1) + \pi_2 p(\mathbf{x} | y = 2)$$

where $p(\mathbf{x} | y = 1)$ and $p(\mathbf{x} | y = 2)$ are the Gaussian densities for each class and $\pi_1$ and $\pi_2$ are their respective prior probabilities.

3.  **Gaussian Distribution for Each Class**

For each class $m \in {1, 2}$, the conditional probability density function is:

$$p(\mathbf{x} | y = m) = \frac{1}{(2\pi) |\boldsymbol{\Sigma}|^{1/2}} \exp\left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_m)^\top \boldsymbol{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu}_m) \right)$$

Thus, $p(\mathbf{x})$ becomes:

$$p(\mathbf{x}) = \frac{\pi_1}{(2\pi) |\boldsymbol{\Sigma}|^{1/2}} \exp\left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_1)^\top \boldsymbol{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu}_1) \right) 
+ \frac{\pi_2}{(2\pi) |\boldsymbol{\Sigma}|^{1/2}} \exp\left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_2)^\top \boldsymbol{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu}_2) \right)$$

Then, $p(\mathbf{x})$ , can be represented as a weighted sum of the two multivariate Gaussian densities, with the weights being the prior probabilities $\pi_1$ and $\pi_2$.

#### 👾 Problem 1.2

::: boxed-text
Given the training data, estimate the parameters $\pi_m,\boldsymbol{\mu}_m$ for $m=1,2$, and $\boldsymbol{\Sigma}$. Predict the labels of the test data and plot them in a scatter plot with different colors to represent the different classes.
:::

The next code estimates $\pi_m,\boldsymbol{\mu}_m$ for $m=1,2$, and $\boldsymbol{\Sigma}$. Predict the labels of the test data and plot them in a scatter plot with different colors to represent the different classes.

```{r}
rm(list=ls()) # Remove variables 
cat("\014") # Clean workspace
set.seed(1234)
suppressMessages(library(caret))
load(file = '/Users/quant/Desktop/Data Science/Primer semestre/Machine Learning - Spring 2024/Computer Lab 4/cancer_data_10features.RData')
train_obs <- createDataPartition(y = cancer_data_10features$diagnosis, p = .80, list = FALSE)
train <- cancer_data_10features[train_obs, 1:3]
test <- cancer_data_10features[-train_obs, 1:3]
# Estimate parameters
n_train <- nrow(train)
ind_train_M <- train$diagnosis == "M"
pi_1 <- mean(train$diagnosis == "M")  
pi_2 <- 1 - pi_1                      
mu_1 <- colMeans(train[ind_train_M, 2:3])   
mu_2 <- colMeans(train[!ind_train_M, 2:3])  
# Shared covariance matrix

# Extract the data as matrices
X_malign <- as.matrix(train[ind_train_M, 2:3])
X_benign <- as.matrix(train[!ind_train_M, 2:3]) 
Sigma <- (t(X_malign - mu_1) %*% (X_malign - mu_1) + t(X_benign - mu_2) %*% (X_benign - mu_2)) / n_train
# Print the parameters
cat("pi_1 (Malign):", pi_1, "\n")
cat("pi_2 (Benign):", pi_2, "\n")
cat("mu_1 (Malign):", mu_1, "\n")
cat("mu_2 (Benign):", mu_2, "\n")
cat("Shared Covariance Matrix Sigma:\n", Sigma, "\n")

# Decision rule for LDA prediction
predict_LDA <- function(x_star, mu_1, mu_2, Sigma, pi_1, pi_2) {
  inv_Sigma <- solve(Sigma)
  
  score_1 <- t(x_star) %*% inv_Sigma %*% mu_1 - 0.5 * t(mu_1) %*% inv_Sigma %*% mu_1 + log(pi_1)
  score_2 <- t(x_star) %*% inv_Sigma %*% mu_2 - 0.5 * t(mu_2) %*% inv_Sigma %*% mu_2 + log(pi_2)
  
  if (score_1 > score_2) {
    return("M")  
  } else {
    return("B") 
  }
}

# Decision rule applied to test data
test_predictions <- apply(test[, 2:3], 1, predict_LDA, mu_1, mu_2, Sigma, pi_1, pi_2)

# Plot 
ind_test_pred_M <- test_predictions == "M"
ind_test_pred_B <- test_predictions == "B"
plot(test[ind_test_pred_M, 2], test[ind_test_pred_M, 3], col = "cornflowerblue", pch = "x", xlab = "Radius", ylab = "Texture", xlim = c(5, 30), ylim = c(5, 40), main = "LDA: Test Predictions")
points(test[ind_test_pred_B, 2], test[ind_test_pred_B, 3], col = "lightcoral", pch = "x")
legend("topleft", legend = c("Malign (predicted)", "Benign (predicted)"), col = c("cornflowerblue", "lightcoral"), pch = "x", cex = 0.70)
```

#### 👾 Problem 1.3

::: boxed-text
Plot the decision bound of the classifier and the predictions of the test data from Problem 1.2 in the same plot.
:::

We can re-write the equation $\mathbf{x}_{\star}^{\top}\boldsymbol{\Sigma}^{-1}(\boldsymbol{\mu}_{1} - \boldsymbol{\mu}_{2})- \frac{1}{2}\boldsymbol{\mu}_{1}^{\top}\boldsymbol{\Sigma}^{-1}\boldsymbol{\mu}_{1} + \frac{1}{2}\boldsymbol{\mu}_{2}^{\top}\boldsymbol{\Sigma}^{-1}\boldsymbol{\mu}_{2}+\log \pi_{1} - \log \pi_{2}=0.$ as something that looks more like the linear equation for the decision boundary $a_1x_1+ a_2x_2 + intercept = 0$ to code it.

The next code plots the decision bound of the classifier and the predictions of the test data from Problem 1.2 in the same plot.

```{r}
# Difference between means
mu_diff <- mu_1 - mu_2

# Inverse covariance
inv_Sigma <- solve(Sigma)

# Decision boundary terms
a <- inv_Sigma[1, 1] * mu_diff[1] + inv_Sigma[1, 2] * mu_diff[2]
b <- inv_Sigma[2, 1] * mu_diff[1] + inv_Sigma[2, 2] * mu_diff[2]
intercept <- log(pi_1 / pi_2) - 0.5 * (t(mu_1) %*% inv_Sigma %*% mu_1 - t(mu_2) %*% inv_Sigma %*% mu_2)

# Grid of radius and corresponding texture
x1_grid <- seq(min(test[, 2]), max(test[, 2]), length.out = 100)
x2_grid <- -(a * x1_grid + intercept) / b

#Plot 
plot(test[ind_test_pred_M, 2], test[ind_test_pred_M, 3], col = "cornflowerblue", pch = "x", 
     xlab = "Radius", ylab = "Texture", 
     xlim = range(test[, 2]), ylim = range(test[, 3]), main = "Test predictions and the decision bound")
points(test[ind_test_pred_B, 2], test[ind_test_pred_B, 3], col = "lightcoral", pch = "x")

# Decision boundary
lines(x1_grid, x2_grid, col = "black", lwd = 2)

legend("topleft", legend = c("Malign (predicted)", "Benign (predicted)", "Decision Boundary"), 
       col = c("cornflowerblue", "lightcoral", "black"), pch = c("x", "x", NA), lwd = c(NA, NA, 2), cex = 0.7)
```

#### 👾 Problem 1.4

::: boxed-text
Fit a logistic regression to the training data using the `glm()` function. Compare the results to the generative model in Problem 1.2. Comment on the results.
:::

The next code fits a logistic regression to the training data using the `glm()` function. Moreover, it compares the results to the generative model in Problem 1.2.

```{r}
# Logistic regression for the training data
log_reg_model <- glm(diagnosis ~ radius + texture, data = train, family = binomial)

# Predict on the test data
log_reg_probs <- predict(log_reg_model, newdata = test, type = "response")

# Convert probabilities to binary classifications
log_reg_predictions <- ifelse(log_reg_probs > 0.5, "M", "B")

# Confusion matrix for logistic regression
confusion_matrix_log_reg <- table(log_reg_predictions, test$diagnosis)

# Accuracy for logistic regression
accuracy_log_reg <- sum(diag(confusion_matrix_log_reg)) / sum(confusion_matrix_log_reg)

# Print
cat("Logistic Regression Confusion Matrix:\n")
print(confusion_matrix_log_reg)
cat("\nLogistic Regression Accuracy:", accuracy_log_reg, "\n")

# Confusion matrix for LDA
confusion_matrix_lda <- table(test_predictions, test$diagnosis)

# Accuracy for LDA
accuracy_lda <- sum(diag(confusion_matrix_lda)) / sum(confusion_matrix_lda)

# Print
cat("LDA Confusion Matrix:\n")
print(confusion_matrix_lda)
cat("\nLDA Accuracy:", accuracy_lda, "\n")
```

Despite having a similar accuracy, the confusion matrices reveal that using a Logistic Regression is more favorable in this situation, where reducing false negatives is crucial for the patients.

## 2. Quadratic discriminant analysis for cancer data

#### 👾 Problem 2.1

::: boxed-text
Derive (analytically) $p(\mathbf{x})$ with the new class conditional distribution above.
:::

As each one of the classes has its own covariance, we can write $p(\mathbf{x})$ as a weighted sum of the two multivariate Gaussian densities, with the weights being the prior probabilities $\pi_1$ and $\pi_2$ and a covariance matrix ${\Sigma}_m$ for each class (similar to Problem 1.1):

$p(\mathbf{x}) = \sum_{m=1}^{2} \pi_m \frac{1}{(2\pi)^{d/2} |\boldsymbol{\Sigma}_m|^{1/2}} \exp\left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_m)^\top \boldsymbol{\Sigma}_m^{-1} (\mathbf{x} - \boldsymbol{\mu}_m) \right)$

#### 👾 Problem 2.2

::: boxed-text
Given the training data, estimate the parameters $\pi_m,\boldsymbol{\mu}_m$, and $\boldsymbol{\Sigma}_m$ for $m=1,2$. Predict the labels of the test data and plot them in a scatter plot with different colors to represent the different classes.
:::

The next code estimates the parameters $\pi_m,\boldsymbol{\mu}_m$, and $\boldsymbol{\Sigma}_m$ for $m=1,2$. Then, predicts the labels of the test data and plot them in a scatter plot with different colors to represent the different classes:

```{r}

# Number of observations per class
n_malign <- nrow(X_malign)
n_benign <- nrow(X_benign)

# Sigma_1 and Sigma_2 (covariance for each class)
Sigma_1 <- t(X_malign - mu_1) %*% (X_malign - mu_1) / n_malign
Sigma_2 <- t(X_benign - mu_2) %*% (X_benign - mu_2) / n_benign

# QDA Prediction function 
predict_QDA <- function(x_star, mu_1, mu_2, Sigma_1, Sigma_2, pi_1, pi_2) {
  inv_Sigma_1 <- solve(Sigma_1)
  inv_Sigma_2 <- solve(Sigma_2)
  
  log_det_Sigma_1 <- log(det(Sigma_1))
  log_det_Sigma_2 <- log(det(Sigma_2))
  
  # Score for Malign
  score_1 <- -0.5 * log_det_Sigma_1 - 0.5 * t(x_star - mu_1) %*% inv_Sigma_1 %*% (x_star - mu_1) + log(pi_1)
  
  # Score for Benign
  score_2 <- -0.5 * log_det_Sigma_2 - 0.5 * t(x_star - mu_2) %*% inv_Sigma_2 %*% (x_star - mu_2) + log(pi_2)
  
  if (score_1 > score_2) {
    return("M")
  } else {
    return("B")
  }
}

# Print the parameters
cat("pi_1 (Malign):", pi_1, "\n")
cat("pi_2 (Benign):", pi_2, "\n")
cat("mu_1 (Malign):", mu_1, "\n")
cat("mu_2 (Benign):", mu_2, "\n")
cat(" Sigma For (Maling) :\n", Sigma_1, "\n")
cat(" Sigma For (Benign) :\n", Sigma_2, "\n")

# QDA decision rule on the test data
test_predictions_qda <- apply(test[, 2:3], 1, predict_QDA, mu_1, mu_2, Sigma_1, Sigma_2, pi_1, pi_2)

# Plot
ind_test_pred_M <- test_predictions_qda == "M"
ind_test_pred_B <- test_predictions_qda == "B"
plot(test[ind_test_pred_M, 2], test[ind_test_pred_M, 3], col = "cornflowerblue", pch = "x", 
     xlab = "Radius", ylab = "Texture", xlim = c(5, 30), ylim = c(5, 40), 
     main = "QDA: Test Predictions")
points(test[ind_test_pred_B, 2], test[ind_test_pred_B, 3], col = "lightcoral", pch = "x")
legend("topleft", legend = c("Malign (predicted)", "Benign (predicted)"), 
       col = c("cornflowerblue", "lightcoral"), pch = "x", cex = 0.70)
```

#### 👾 Problem 2.3

::: boxed-text
Compare the quadratic discriminant classifier to the linear discriminant and logistic classifiers from Problem 1. Discuss the results.
:::

The next code compares the quadratic discriminant classifier to the linear discriminant and logistic classifiers from Problem 1:

```{r}
# Confusion matrix and accuracy for QDA
confusion_matrix_qda <- table(test_predictions_qda, test$diagnosis)
accuracy_qda <- sum(diag(confusion_matrix_qda)) / sum(confusion_matrix_qda)

# Print
cat("QDA Confusion Matrix:\n")
print(confusion_matrix_qda)
cat("\nQDA Accuracy:", accuracy_qda, "\n")
cat("LDA Confusion Matrix:\n")
print(confusion_matrix_lda)
cat("\nLDA Accuracy:", accuracy_lda, "\n")
cat("Logistic Regression Confusion Matrix:\n")
print(confusion_matrix_log_reg)
cat("\nLogistic Regression Accuracy:", accuracy_log_reg, "\n")

```

The confusion matrices reveal that, as in Problem 1.4, using a Logistic Regression is more favorable in this situation, where reducing false negatives is crucial for the patients.

#### 👾 Problem 2.4

::: boxed-text
A doctor contacts you and says she has a patient whose digitised image has the features `radius=13.20` and `texture=19.22`. Use your best classifier from Problem 2.3 to provide the doctor with some advice.
:::

The next code uses the logistic regression model to predict if the digitised image of a patient with features `radius=13.20` and `texture=19.22` is benign or malign:

```{r}
# Patient's data
patient_data <- data.frame(radius = 13.20, texture = 19.22)
patient_prediction <- predict(log_reg_model, newdata = patient_data, type = "response")

cat("Predicted diagnosis probability (malign) for the patient with Radius =", patient_data$radius, "and Texture =", patient_data$texture, "is:", patient_prediction, "\n")

predicted_class <- ifelse(patient_prediction > 0.5, "M", "B")
cat("The predicted diagnosis is:", predicted_class, "\n")
```

Doctor X, based on the patient's digitized image features, our best classifier indicates that the tumor is likely to be benign. However, I recommend monitoring the patient and conducting regular check-ups to ensure the diagnosis remains accurate.

## 3. Unsupervised Gaussian mixture models

#### 👾 Problem 3.1

::: boxed-text
Give a possible interpretation of the two clusters.
:::

In the context of the stock data example, the two clusters reflect two different market regimes:

Cluster 1 represents periods of low activity (bear market), where trading volumes are lower, and stock prices are depressed.

Cluster 2 represents periods of higher market activity (bull market), where trading volumes are higher, and stock prices are elevated.

#### 👾 Problem 3.2

::: boxed-text
Derive (analytically) $p(x)$ for the example above.
:::

$p(x)$ is a weighted sum of the class-conditional probabilities, weighted by the class priors $( \pi_1 )$ and $( \pi_2 )$: $p(x) = \pi_1 p(x | y = 1) + \pi_2 p(x | y = 2)$

As each class follows a normal distribution, $x | y = m \sim \mathcal{N}(\mu_m, \sigma_m^2), \, m = 1, 2$, this means that $p(x | y = 1) = \frac{1}{\sqrt{2\pi\sigma_1^2}} \exp \left( -\frac{(x - \mu_1)^2}{2\sigma_1^2} \right)$, and $p(x | y = 2) = \frac{1}{\sqrt{2\pi\sigma_2^2}} \exp \left( -\frac{(x - \mu_2)^2}{2\sigma_2^2} \right)$.

Given that, $\mu_1 = -2, \sigma_1 = 0.5$ for class 1, and $\mu_2 = 4, \sigma_2 = 1.5$, for class 2, and the probabilities priors $\pi_1 = 0.2$ and $\pi_2 = 0.8$, we can plug these values:

$p(x) = 0.2 \cdot \frac{1}{\sqrt{2\pi(0.5^2)}} \exp \left( -\frac{(x + 2)^2}{2(0.5^2)} \right) + 0.8 \cdot \frac{1}{\sqrt{2\pi(1.5^2)}} \exp \left( -\frac{(x - 4)^2}{2(1.5^2)} \right)$

Simplifying the constants:

$p(x) = 0.2 \cdot \frac{1}{\sqrt{\pi}} \exp \left( -\frac{(x + 2)^2}{0.5^2} \right) + 0.8 \cdot \frac{1}{\sqrt{4.5\pi}} \exp \left( -\frac{(x - 4)^2}{4.5} \right)$

#### 👾 Problem 3.3

::: boxed-text
Plot the $p(x)$ you derived in Problem 3.2 in the same figure as a (normalised) histogram.
:::

```{r}
# Parameters 
mu1 <- -2
sigma1 <- 0.5
mu2 <- 4
sigma2 <- 1.5
pi1 <- 0.2
pi2 <- 1 - pi1

# Store in vectors
mu <- c(mu1, mu2)
sigma <- c(sigma1, sigma2)
pis <- c(pi1, pi2)
n <- 1000
y <- rep(NA, n)
x <- rep(NA, n)

# Simulate data from the Gaussian mixture model
set.seed(1234)  # For reproducibility
for(i in 1:n){
  # Simulate indicator with probability pi2 (1 - component 2, 0 - component 1)
  y[i] <- rbinom(n = 1, size=1, prob = pis[2]) + 1
  x[i] <- rnorm(n = 1, mean = mu[y[i]], sd = sigma[y[i]])
}

# Sanity check
cat("Estimated class 1 marginal probability: ", mean(y == 1), ". True: ", pis[1], sep = "\n")
cat("Estimated class 2 marginal probability: ", mean(y == 2), ". True: ", pis[2], sep = "\n")

# Histogram of the simulated data
hist(x, breaks = 30, prob = TRUE, xlab = "x", col = "cornflowerblue", main = "Normalized histogram of simulated x and theoretical p(x)")

# Grid of x values
x_grid <- seq(-6, 10, length.out = 1000)

# p(x) from Problem 3.2
p_x <- pi1 * (1 / (sqrt(2 * pi * sigma1^2))) * exp(-((x_grid - mu1)^2) / (2 * sigma1^2)) + 
       pi2 * (1 / (sqrt(2 * pi * sigma2^2))) * exp(-((x_grid - mu2)^2) / (2 * sigma2^2))

lines(x_grid, p_x, col = "lightcoral", lwd = 2)

legend("topright", legend = c("Simulated Data", "Theoretical p(x)"), 
       fill = c("cornflowerblue", NA), lty = c(NA, 1), lwd = c(NA, 2), 
       border = c("black", NA), col = c(NA, "lightcoral"), cex = 0.7)
```

## 4. Unsupervised learning via the EM algorithm

```{r}
EM_GMM_M2 <- function(x, mu_start, sigma_start, pis_start, n_iter = 100) {
  # Estimates the parameters in an unsupervised Gaussian mixture model with M = 2 classes. 
  # Runs the EM algorithm for n_iter iterations. x is assumed univariate. 
  # mu_start, sigma_start, pis_start are starting values.
  stopifnot(sum(pis_start) == 1)
  # Quantities to save
  pis_all <-  matrix(NA, nrow = n_iter, ncol = 2)
  mu_all <- matrix(NA, nrow = n_iter, ncol = 2)
  sigma_all <- matrix(NA, nrow = n_iter, ncol = 2)
  Q_all <- rep(NA, n_iter)
  log_like_all <- rep(NA, n_iter)
  
  # Initialise
  mu <- mu_start
  sigma <- sigma_start
  pis <- pis_start
  n <- length(x)
  W <- matrix(0, nrow = n, ncol = 2) # Unnormalised weights for each observation
  log_pdf_class <- matrix(0, nrow = n, ncol = 2) # The log-likelihood of the two classes for each obs. To compute Q. 
  for(j in 1:n_iter){
    # Start EM steps
    # E-step: Compute the expected log-likelihood Q
    for(m in 1:2){
      # The log-density for each class
      log_pdf_class[, m] <- dnorm(x, mean = mu[m], sd = sigma[m], log = TRUE) + log(pis[m])
      # Unnormalised weights
      W[, m] <- pis[m]*dnorm(x, mean = mu[m], sd = sigma[m])
    }
    w <- W/rowSums(W) # Normalise weights
    n_hat <- colSums(w) # Expected number of obs per class
    Q <- sum(rowSums(w*log_pdf_class)) # Expected log-likelihood
    
    # M-step: Maximise Q. Closed form analytical solution in Gaussian mixture models
    for(m in 1:2){
      pis[m] <- n_hat[m]/n
      mu[m] <- 1/n_hat[m]*sum(w[, m]*x)
      sigma[m] <- sqrt(1/n_hat[m]*sum(w[, m]*(x - mu[m])^2))
    }
    # End EM steps. Save estimates, Q, and log-likelihood
    pis_all[j, ] <- pis
    mu_all[j, ] <- mu
    sigma_all[j, ] <- sigma
    Q_all[j] <- Q
    # Compute log-likelihood at current parameter values
    for(m in 1:2){
      # Unnormalised weights
      W[, m] <- pis[m]*dnorm(x, mean = mu[m], sd = sigma[m])
    }
    log_like_all[j] <-  sum(log(rowSums(W)))
  } # End EM iterations
  # Return everything as a list
  return(list(pi_hat = pis, mu_hat = mu, sigma_hat = sigma, 
              weights = W/rowSums(W), pis_all = pis_all, 
              mu_all = mu_all, sigma_all = sigma_all, Q_all = Q_all, 
              log_like_all = log_like_all))}
```

```{r}
# Initial values
pis_start <- c(0.5, 0.5)
mu_start <- c(1, 4)
sigma_start <- c(1, 3)
n_iter <- 20
EM_result <- EM_GMM_M2(x, mu_start, sigma_start, pis_start, n_iter = n_iter)

# Visualise convergence for each parameters (adding starting value)
matplot(0:n_iter, rbind(pis_start, EM_result$pis_all), main = 'pis', pch = c("o", "o"), col = c("cornflowerblue", "lightcoral"), xlab = "Iteration", ylab = "pis", ylim = c(0, 1.5))
legend("topright", legend = c("Class 1", "Class 2"), col = c("cornflowerblue", "lightcoral"), pch = c("o", "o"))
matplot(0:n_iter, rbind(mu_start, EM_result$mu_all), main = 'mu', pch = c("o", "o"), col = c("cornflowerblue", "lightcoral"), xlab = "Iteration", ylab = "mu", ylim = c(-3, 6))
legend("topright", legend = c("Class 1", "Class 2"), col = c("cornflowerblue", "lightcoral"), pch = c("o", "o"))
matplot(0:n_iter, rbind(sigma_start, EM_result$sigma_all), main = 'sigma', pch = c("o", "o"), col = c("cornflowerblue", "lightcoral"), xlab = "Iteration", ylab = "sigma", ylim = c(0, 4))
legend("topright", legend = c("Class 1", "Class 2"), col = c("cornflowerblue", "lightcoral"), pch = c("o", "o"))

par(mfrow = c(1, 1))
# Inspect convergence
plot(EM_result$log_like_all, main = 'Log-likelihood', type = "l", col = "cornflowerblue", xlab = "Iteration", ylab = "Log-likelihood")
plot(EM_result$Q_all, main = 'Expected log-likelihood (Q)', type = "l", col = "cornflowerblue", xlab = "Iteration", ylab = "Log-likelihood")

print("True parameters (pi, mu, and sigma)")
print(pis)
print(mu)
print(sigma)
print("Estimated parameters (pi, mu, and sigma)")
print(EM_result$pi_hat)
print(EM_result$mu_hat)
print(EM_result$sigma_hat)

# Inspect the classification probabilities of observation 10
ind <- 30
barplot(names.arg = c("Class 1", "Class 2"), EM_result$weights[ind, ], col = "cornflowerblue", ylim = c(0, 1), main = paste("Class (posterior) probability observation ", ind, sep = ""))
```

#### 👾 Problem 4.1

::: boxed-text
Explain the label-switching problem when estimating unsupervised Gaussian mixture models. Do you observe label-switching above?
:::

In unsupervised GMMs, the EM algorithm can swap labels between classes since the labels are interchangeable, causing confusion but not affecting the model's outcome.

In above plots, there is no evidence of label-switching, as the parameters (pis, mu) remain stable across iterations for both classes, but sigma crossed around the 10th iteration so it might be indicating temporary label-switching but converged smoothly afterward.

#### 👾 Problem 4.2

::: boxed-text
Use the `EM_GMM_M2()` function to estimate an unsupervised Gaussian mixture model for the insect data in Problem 3. Analyse the convergence. Compare the parameter estimates to those obtained by the function `normalmixEM()` in the `mixtools` package.
:::

```{r}
# Load Data
load(file = '/Users/quant/Desktop/Data Science/Primer semestre/Machine Learning - Spring 2024/Computer Lab 4/insects.RData')
hist(insects$length, col = "cornflowerblue", main = "Histogram of insects' lengths", prob = TRUE, xlab = "Length", ylim = c(0, 0.4), xlim = c(0, 14))
abline(v = insects[6, ], lwd = 1.5, col = "lightcoral")
abline(v = insects[244, ], lwd = 1.5, col = "purple")
abline(v = insects[421, ], lwd = 1.5, col = "lightpink")
legend("topright", legend = c("Obs 6", "Obs 244", "Obs 421"), col = c("lightcoral", "purple", "lightpink"), lwd = c(1.5, 1.5, 1.5))

# Initial values
pis_start <- c(0.4, 0.6)
mu_start <- c(6, 3)
sigma_start <- c(0.5, 0.6)
n_iter <- 30
EM_result <- EM_GMM_M2(insects$length, mu_start, sigma_start, pis_start, n_iter = n_iter)

# Visualise convergence for each parameters (adding starting value)
matplot(0:n_iter, rbind(pis_start, EM_result$pis_all), main = 'pis', pch = c("o", "o"), col = c("cornflowerblue", "lightcoral"), xlab = "Iteration", ylab = "pis", ylim = c(0, 1.5))
legend("topright", legend = c("Class 1", "Class 2"), col = c("cornflowerblue", "lightcoral"), pch = c("o", "o"))
matplot(0:n_iter, rbind(mu_start, EM_result$mu_all), main = 'mu', pch = c("o", "o"), col = c("cornflowerblue", "lightcoral"), xlab = "Iteration", ylab = "mu", ylim = c(-3, 12))
legend("topright", legend = c("Class 1", "Class 2"), col = c("cornflowerblue", "lightcoral"), pch = c("o", "o"))
matplot(0:n_iter, rbind(sigma_start, EM_result$sigma_all), main = 'sigma', pch = c("o", "o"), col = c("cornflowerblue", "lightcoral"), xlab = "Iteration", ylab = "sigma", ylim = c(0, 4))
legend("topright", legend = c("Class 1", "Class 2"), col = c("cornflowerblue", "lightcoral"), pch = c("o", "o"))

par(mfrow = c(1, 1))
# Inspect convergence
plot(EM_result$log_like_all, main = 'Log-likelihood', type = "l", col = "cornflowerblue", xlab = "Iteration", ylab = "Log-likelihood")
plot(EM_result$Q_all, main = 'Expected log-likelihood (Q)', type = "l", col = "cornflowerblue", xlab = "Iteration", ylab = "Log-likelihood")

print("Estimated parameters (pi, mu, and sigma)")
print(EM_result$pi_hat)
print(EM_result$mu_hat)
print(EM_result$sigma_hat)
```

The Gaussian mixture model using EM_GMM_M2() function algorithm had converged around iteration 15, as seen from the stable log-likelihood, pis, mu and sigma we also tried different starting values, and this result gave the highest log-likelihood.

```{r}

library(mixtools)
normalmix_insects <- normalmixEM(insects$length, k = 2) # k = 2 for two components

print("Estimated parameters From EM_GMM_M2() (pi, mu, and sigma) ")
print(EM_result$pi_hat)
print(EM_result$mu_hat)
print(EM_result$sigma_hat)

print("Estimated parameters From normalmixE (pi, mu, and sigma)")
print(normalmix_insects$lambda) # Mixing coefficients
print(normalmix_insects$mu)     # Means
print(normalmix_insects$sigma)  # Standard deviations
```

Both models estimated the parameters very similarly.

#### 👾 Problem 4.3

::: boxed-text
Plot the class posterior probabilities for insects 6, 244, and 421 with the model estimated in Problem 4.2. Are the results as expected? Explain.
:::

```{r}
ind <- 6
barplot(names.arg = c("Insect 1", "Insect 2"), EM_result$weights[ind, ], col = "cornflowerblue", ylim = c(0, 1), main = paste("Class (posterior) probability observation ", ind, 
                     "\nLength = ", round(insects$length[ind],3),  sep = ""))

ind <- 244
barplot(names.arg = c("Insect 1", "Insect 2"), EM_result$weights[ind, ], col = "cornflowerblue", ylim = c(0, 1), main = paste("Class (posterior) probability observation ", ind, "\nLength = ", round(insects$length[ind],3),sep = ""))

ind <- 421
barplot(names.arg = c("Insect 1", "Insect 2"), EM_result$weights[ind, ], col = "cornflowerblue", ylim = c(0, 1), main = paste("Class (posterior) probability observation ", ind,"\nLength = ", round(insects$length[ind],3), sep = ""))

```

Observation 6: For observation 6, the posterior probability is fully assigned to Insect 2. This makes sense since its length is on the lower end, closer to the mean of Insect 2, as shown in the histogram.

Observation 244: The posterior probability for observation 244 is split, with a higher chance for Insect 1 but a noticeable portion still going to Insect 2. This matches its location in the middle, where the two distributions overlap.

Observation 421: For observation 421, the posterior probability is entirely given to Insect 1. This aligns with its length being on the higher side.

#### 👾 Problem 4.4

::: boxed-text
Write a function that implements the EM algorithm for the unsupervised Gaussian mixture model with any number of components $M$, i.e. not limited to $M=2$ as the `EM_GMM_M2()` function. You can still assume that $x$ is univariate. Use your function to estimate two models for the dataset `fish.Rdata` with, respectively, $M=3$ and $M=4$ classes. The dataset can be downloaded from the Canvas page of the course. The dataset contains the lengths of 523 fishes.
:::

```{r}
EM_GMM <- function(x, mu_start, sigma_start, pis_start, n_iter = 100, M) {
  # Estimates the parameters in an unsupervised Gaussian mixture model with M = 2 classes. 
  # Runs the EM algorithm for n_iter iterations. x is assumed univariate. 
  # mu_start, sigma_start, pis_start are starting values.
  stopifnot(sum(pis_start) == 1)
  # Quantities to save
  pis_all <-  matrix(NA, nrow = n_iter, ncol = M)
  mu_all <- matrix(NA, nrow = n_iter, ncol = M)
  sigma_all <- matrix(NA, nrow = n_iter, ncol = M)
  Q_all <- rep(NA, n_iter)
  log_like_all <- rep(NA, n_iter)
  
  # Initialise
  mu <- mu_start
  sigma <- sigma_start
  pis <- pis_start
  n <- length(x)
  W <- matrix(0, nrow = n, ncol = M) # Unnormalised weights for each observation
  log_pdf_class <- matrix(0, nrow = n, ncol = M) # The log-likelihood of the two classes for each obs. To compute Q. 
  for(j in 1:n_iter){
    # Start EM steps
    # E-step: Compute the expected log-likelihood Q
    for(m in 1:M){
      # The log-density for each class
      log_pdf_class[, m] <- dnorm(x, mean = mu[m], sd = sigma[m], log = TRUE) + log(pis[m])
      # Unnormalised weights
      W[, m] <- pis[m]*dnorm(x, mean = mu[m], sd = sigma[m])
    }
    w <- W/rowSums(W) # Normalise weights
    n_hat <- colSums(w) # Expected number of obs per class
    Q <- sum(rowSums(w*log_pdf_class)) # Expected log-likelihood
    
    # M-step: Maximise Q. Closed form analytical solution in Gaussian mixture models
    for(m in 1:M){
      pis[m] <- n_hat[m]/n
      mu[m] <- 1/n_hat[m]*sum(w[, m]*x)
      sigma[m] <- sqrt(1/n_hat[m]*sum(w[, m]*(x - mu[m])^2))
    }
    # End EM steps. Save estimates, Q, and log-likelihood
    pis_all[j, ] <- pis
    mu_all[j, ] <- mu
    sigma_all[j, ] <- sigma
    Q_all[j] <- Q
    # Compute log-likelihood at current parameter values
    for(m in 1:M){
      # Unnormalised weights
      W[, m] <- pis[m]*dnorm(x, mean = mu[m], sd = sigma[m])
    }
    log_like_all[j] <-  sum(log(rowSums(W)))
  } # End EM iterations
  # Return everything as a list
  return(list(pi_hat = pis, mu_hat = mu, sigma_hat = sigma, 
              weights = W/rowSums(W), pis_all = pis_all, 
              mu_all = mu_all, sigma_all = sigma_all, Q_all = Q_all, 
              log_like_all = log_like_all))
}
```

```{r}
# Load Data
load(file = '/Users/quant/Desktop/Data Science/Primer semestre/Machine Learning - Spring 2024/Computer Lab 4/fish.RData')

# Plot the histogram with 30 breaks
hist(fish$length, breaks = 30, xlab = "Length (in mm)", ylab = "Counts", 
     col = "cornflowerblue", main = "Histogram of the lengths of Fish")
```

```{r}
# For M = 3
# Initial values
pis_start <- c(0.3, 0.3,0.4)
mu_start <- c(10, 20,30)
sigma_start <- c(5, 6,7)
n_iter <- 100
EM_result_fish_3 <- EM_GMM(fish$length, mu_start, sigma_start, pis_start, n_iter = n_iter,M=3)

print("Estimated parameters (pi, mu, and sigma)")
print(EM_result_fish_3$pi_hat)
print(EM_result_fish_3$mu_hat)
print(EM_result_fish_3$sigma_hat)
```

```{r}
# For M = 4
# Initial values
pis_start <- c(0.25, 0.25,0.25,0.25)
mu_start <- c(20, 30,45,60)
sigma_start <- c(5, 8,8,8)
n_iter <- 100
EM_result_fish_4 <- EM_GMM(fish$length, mu_start, sigma_start, pis_start, n_iter = n_iter,M=4)

print("Estimated parameters (pi, mu, and sigma)")
print(EM_result_fish_4$pi_hat)
print(EM_result_fish_4$mu_hat)
print(EM_result_fish_4$sigma_hat)
```

#### 👾 Problem 4.5

::: boxed-text
Plot $p(x)$ (using the estimates from your EM algorithm) for the two models in Problem 4.4 in the same figure as a (normalised) histogram obtained with the `hist()` with the argument `breaks=30`. Which of the two models seem visually better for modelling the fishs' lengths?
:::

```{r}
# Load data
#load(file = './fish.RData')
x <- fish$length

# Function to calculate the GMM density
gmm_density <- function(x, mu, sigma, pis) {
  M <- length(mu)
  density <- rep(0, length(x))
  for (m in 1:M) {
    density <- density + pis[m] * dnorm(x, mean = mu[m], sd = sigma[m])
  }
  return(density)
}

# Plot the normalized histogram
hist(x, breaks = 30, probability = TRUE, col = "cornflowerblue", 
     xlab = "Length (in mm)", main = "Fish Lengths and GMM Fit")


# Use the results from your EM_GMM for 4 components
mu_3 <- EM_result_fish_3$mu_hat
sigma_3 <- EM_result_fish_3$sigma_hat
pis_3 <- EM_result_fish_3$pi_hat
curve(gmm_density(x, mu_3, sigma_3, pis_3), add = TRUE, col = "red", lwd = 2,lty=2)

# Use the results from your EM_GMM for 4 components
mu_4 <- EM_result_fish_4$mu_hat
sigma_4 <- EM_result_fish_4$sigma_hat
pis_4 <- EM_result_fish_4$pi_hat
curve(gmm_density(x, mu_4, sigma_4, pis_4), add = TRUE, col = "black", lwd = 2)

# Add legend
legend("topright", legend = c("3-component GMM", "4-component GMM"), 
       col = c("red", "black"), lty = c(2, 1), lwd = 2)
```

The 3-component Gaussian mixture model (black line) seems to provide a better fit for the fish lengths. It captures the key features more accurately than 4--component Gaussian mixture model.

#### 👾 Problem 4.6

::: boxed-text
Use the function `mvnormalmixEM()` from the `mixtools` package to estimate an unsupervised (multivariate) Gaussian mixture model with $M=2$ classes to the ASB stock example with feature vector $\mathbf{x}=(x_1,x_2)^\top$, with $x_1=\text{Close}$ and $x_2=\text{log(Volume)}$. Use all observations as training data. Plot a scatter plot with the predicted classes for the training data (use different colors to represent the different classes).
:::

```{r}
load(file = '/Users/quant/Desktop/Data Science/Primer semestre/Machine Learning - Spring 2024/Computer Lab 4/asb.RData')


# Transform Volume to Log Volume
asb$Log_Volume <- log(asb$Volume)
data = asb[ , c("Close", "Log_Volume")]

# Load necessary libraries
library(mixtools)
library(ggplot2)

# Fit a 2-component Gaussian mixture model (GMM)
gmm_result <- mvnormalmixEM(data, k = 2)

# Get the predicted class for each observation (choose the class with the highest posterior probability)
predicted_classes <- apply(gmm_result$posterior, 1, which.max)

# Add the predicted classes to your original dataset
data$predicted_class <- predicted_classes

# Plot the scatter plot, using different colors for the classes
ggplot(data, aes(x = Log_Volume, y = Close, color = factor(predicted_class))) +
  geom_point(size = 2) +
  labs(title = "Scatter Plot of ASB Stock Data with Predicted Classes",
       x = "Log (Volume)",
       y = "Close Price",
       color = "Predicted Class") +
  theme_minimal()

```

We can see in the above plot that Gaussian mixture model effectively separates the two classes.

## 5. Semi supervised learning

#### 👾 Problem 5.1

::: boxed-text
Estimate a (fully) supervised Gaussian mixture model for the penguin data with species as labels and a single feature flipper length. Use separate means and variances for the classes.
:::

```{r}
load(file = '/Users/quant/Desktop/Data Science/Primer semestre/Machine Learning - Spring 2024/Computer Lab 4/palmer_penguins_missing.RData')
# Remove Missing Data
df_no_missing <- na.omit(palmer_penguins_missing)

# Selected only flipper_length_mm : Feature , species : Labels
df_no_missing <- df_no_missing[, c("species", "flipper_length_mm")]

# Split the data by species
adelie_df <- df_no_missing[df_no_missing$species == "Adelie", "flipper_length_mm"]
gentoo_df <- df_no_missing[df_no_missing$species == "Gentoo", "flipper_length_mm"]

```

```{r}
# QDA Prediction function 
predict_QDA_penguins <- function(x_star, mu_1, mu_2, Sigma_1, Sigma_2, pi_1, pi_2) {
 
  inv_Sigma_1 <- solve(Sigma_1)
  inv_Sigma_2 <- solve(Sigma_2)
  
  log_det_Sigma_1 <- log(det(Sigma_1))
  log_det_Sigma_2 <- log(det(Sigma_2))
  
  # Score for Adelie
  score_1 <- -0.5 * log_det_Sigma_1 - 0.5 * t(x_star - mu_1) %*% inv_Sigma_1 %*% (x_star - mu_1) + log(pi_1)
  
  # Score for Gentoo
  score_2 <- -0.5 * log_det_Sigma_2 - 0.5 * t(x_star - mu_2) %*% inv_Sigma_2 %*% (x_star - mu_2) + log(pi_2)
  
  if (score_1 > score_2) {
    return("Adelie")
  } else {
    return("Gentoo")
  }
}
```

```{r}
# Means and variance matrices for Adelie and Gentoo
mu_adelie <-  matrix(mean((adelie_df)))
mu_gentoo <- matrix(mean((gentoo_df)))

# Number of observations per class
n_adelie <- dim(matrix(adelie_df))[1]
n_gentoo <- dim(matrix(gentoo_df))[1]

# Sigma_1 and Sigma_2 (covariance for each class)
Sigma_adelie <- t(adelie_df - mu_adelie) %*% (adelie_df - mu_adelie) / n_adelie
Sigma_gentoo <- t(gentoo_df - mu_gentoo) %*% (gentoo_df - mu_gentoo) / n_gentoo

# Priors
pi_adelie <- mean(df_no_missing$species == "Adelie") 
pi_gentoo <- 1 - pi_adelie 

# Display the estimated parameters
cat("Adelie: Mean =", mu_adelie, "Variance =", Sigma_adelie, "\n")
cat("Gentoo: Mean =", mu_gentoo, "Variance =", Sigma_gentoo, "\n")

# Sample observation
x_star <- c(180)
# Prediction Using QDA
prediction <- predict_QDA_penguins(x_star, mu_adelie, mu_gentoo, Sigma_adelie, Sigma_gentoo, pi_adelie, pi_gentoo)
print(paste("Predicted class:", prediction))
```

#### 👾 Problem 5.2

::: boxed-text
Estimate a semi supervised Gaussian mixture model for the penguin data with species as labels and a single feature flipper length. Compare the parameter estimates to those in Problem 5.1. Comment on the result.

As discussed in the lecture, semi supervised learning can be cast as a special version of the EM algorithm. Adapt the `EM_GMM_M2` function to take an extra argument that contains a vector of labels (with `NA` for unknown labels) and construct the weights accordingly. Note that the log-likelihood of the non-missing observations is monitored (for convergence) in semi supervised learning. This requires modifying `log_like_all`.
:::

```{r}
# EM for semi-supervised 
EM_GMM_M2_semi <- function(x_unlabeled, x_labeled, y_labeled, mu_start, sigma_start, pis_start, n_iter = 100) {
  # Estimates parameters in a semi-supervised Gaussian mixture model with M = 2 classes.
  # Uses EM algorithm for n_iter iterations. Assumes univariate data.
  
  stopifnot(sum(pis_start) == 1)

  # Quantities to save
  pis_all <- matrix(NA, nrow = n_iter, ncol = 2)
  mu_all <- matrix(NA, nrow = n_iter, ncol = 2)
  sigma_all <- matrix(NA, nrow = n_iter, ncol = 2)
  Q_all <- rep(NA, n_iter)
  log_like_all <- rep(NA, n_iter)

  # Initialize
  mu <- mu_start
  sigma <- sigma_start
  pis <- pis_start
  n_unlabeled <- length(x_unlabeled)
  n_labeled <- length(x_labeled)
  
  # Combine labeled and unlabeled data
  x_all <- c(x_unlabeled, x_labeled)
  n_total <- length(x_all)
  
  W <- matrix(0, nrow = n_unlabeled, ncol = 2) # Weights for unlabeled data
  log_pdf_class <- matrix(0, nrow = n_unlabeled, ncol = 2) # Log-likelihood for each class for each unlabeled observation

  for(j in 1:n_iter){
    # Start EM steps

    ## E-Step
    # For unlabeled data
    for(m in 1:2){
      # The log-density for each class
      log_pdf_class[, m] <- dnorm(x_unlabeled, mean = mu[m], sd = sigma[m], log = TRUE) + log(pis[m])
      # Unnormalized weights
      W[, m] <- pis[m] * dnorm(x_unlabeled, mean = mu[m], sd = sigma[m])
    }
    w_unlabeled <- W / rowSums(W) # Normalized weights for unlabeled data
    n_hat_unlabeled <- colSums(w_unlabeled) # Expected number of observations per class from unlabeled data

    # For labeled data, directly assign weights based on known labels
    w_labeled <- matrix(0, nrow = n_labeled, ncol = 2)
    for(i in 1:n_labeled) {
      w_labeled[i, y_labeled[i]] <- 1 # Direct assignment based on known label
    }
    n_hat_labeled <- colSums(w_labeled) # Count of observations in each class from labeled data

    ## M-Step: Update parameters based on both labeled and unlabeled data
    for(m in 1:2){
      # Mixing proportion
      pis[m] <- (n_hat_unlabeled[m] + n_hat_labeled[m]) / n_total

      # Mean update, including contributions from labeled data
      mu[m] <- (sum(w_unlabeled[, m] * x_unlabeled) + sum(w_labeled[, m] * x_labeled)) / 
               (n_hat_unlabeled[m] + n_hat_labeled[m])

      # Standard deviation update, including contributions from labeled data
      sigma[m] <- sqrt((sum(w_unlabeled[, m] * (x_unlabeled - mu[m])^2) + 
                        sum(w_labeled[, m] * (x_labeled - mu[m])^2)) / 
                       (n_hat_unlabeled[m] + n_hat_labeled[m]))
    }
    
    # End EM steps. Save estimates, Q, and log-likelihood
    pis_all[j, ] <- pis
    mu_all[j, ] <- mu
    sigma_all[j, ] <- sigma
    Q_all[j] <- sum(rowSums(w_unlabeled * log_pdf_class)) # Only include unlabeled data for Q calculation

    # Compute log-likelihood at current parameter values (includes all data)
    log_like_all[j] <- sum(log(rowSums(W))) + sum(dnorm(x_labeled, mean = mu[y_labeled], sd = sigma[y_labeled], log = TRUE))
  } # End EM iterations

  # Return everything as a list
  return(list(pi_hat = pis, mu_hat = mu, sigma_hat = sigma, 
              weights = rbind(w_unlabeled, w_labeled), pis_all = pis_all, 
              mu_all = mu_all, sigma_all = sigma_all, Q_all = Q_all, 
              log_like_all = log_like_all))
}
```

```{r}
# Separate labeled and unlabeled data
labeled_data <- palmer_penguins_missing[!is.na(palmer_penguins_missing$species), ]
unlabeled_data <- palmer_penguins_missing[is.na(palmer_penguins_missing$species), ]

x_labeled <- labeled_data$flipper_length
y_labeled <- as.numeric(as.factor(labeled_data$species))  # Convert species to numeric labels

x_unlabeled <- unlabeled_data$flipper_length

# Initial guesses for parameters based on labeled data
initial_mu <- tapply(x_labeled, y_labeled, mean, na.rm = TRUE)
initial_sigma <- tapply(x_labeled, y_labeled, sd, na.rm = TRUE)
initial_pis <- table(y_labeled) / length(y_labeled)

# Convert initial parameters to vectors
mu_start <- as.numeric(initial_mu)
sigma_start <- as.numeric(initial_sigma)
pis_start <- as.numeric(initial_pis)


# Run the semi-supervised EM algorithm
EM_result_penguins <- EM_GMM_M2_semi(x_unlabeled, x_labeled, y_labeled, mu_start, sigma_start, pis_start, n_iter = 50)
```

```{r}
n_iter = 50
matplot(0:n_iter, rbind(pis_start, EM_result_penguins$pis_all), main = 'pis', pch = c("o", "o"), col = c("cornflowerblue", "lightcoral"), xlab = "Iteration", ylab = "pis", ylim = c(0, 1.5))
legend("topright", legend = c("Class 1", "Class 2"), col = c("cornflowerblue", "lightcoral"), pch = c("o", "o"))
matplot(0:n_iter, rbind(mu_start, EM_result_penguins$mu_all), main = 'mu', pch = c("o", "o"), col = c("cornflowerblue", "lightcoral"), xlab = "Iteration", ylab = "mu", ylim = c(100, 250))
legend("topright", legend = c("Class 1", "Class 2"), col = c("cornflowerblue", "lightcoral"), pch = c("o", "o"))
matplot(0:n_iter, rbind(sigma_start, EM_result_penguins$sigma_all), main = 'sigma', pch = c("o", "o"), col = c("cornflowerblue", "lightcoral"), xlab = "Iteration", ylab = "sigma", ylim = c(5, 26))
legend("topright", legend = c("Class 1", "Class 2"), col = c("cornflowerblue", "lightcoral"), pch = c("o", "o"))

par(mfrow = c(1, 1))
# Inspect convergence
plot(EM_result_penguins$log_like_all, main = 'Log-likelihood', type = "l", col = "cornflowerblue", xlab = "Iteration", ylab = "Log-likelihood")
plot(EM_result_penguins$Q_all, main = 'Expected log-likelihood (Q)', type = "l", col = "cornflowerblue", xlab = "Iteration", ylab = "Log-likelihood")

# Display final estimates
cat("Final Mixing Proportions (pis):", EM_result_penguins$pi_hat, "\n")
cat("Final Means (mu):", EM_result_penguins$mu_hat, "\n")
cat("Final Standard Deviations (sigma):", EM_result_penguins$sigma_hat, "\n")
```

#### 👾 Problem 5.3

::: boxed-text
The dataset `palmer_penguins.Rdata`, which can be downloaded from the Canvas page of the course, contains the true values for the labels that are missing in `palmer_penguins_missing.Rdata`. Compute the accuracy of the predictions for these observations using your semi supervised Gaussian mixture model in Problem 5.2. Compare that to the classification obtained via your supervised Gaussian mixture model in Problem 5.1. Comment on the result.
:::

```{r}
load(file = '/Users/quant/Desktop/Data Science/Primer semestre/Machine Learning - Spring 2024/Computer Lab 4/palmer_penguins.RData') 
```

We are going to use NA index from Missing dataframe to use it as out test data from non missing data

```{r}
na_indices <- which(is.na(palmer_penguins_missing$species))
test_data = palmer_penguins[, c("species", "flipper_length_mm")][na_indices,]
```

```{r}
# Prediction from Supervised learning
test_data$pred_supervised <-  sapply(test_data$flipper_length_mm, function(x) {
  predict_QDA_penguins(x, mu_adelie, mu_gentoo, Sigma_adelie, Sigma_gentoo, pi_adelie, pi_gentoo)
})

# Prediction From Unsupervised learning
test_data$pred_semi_supervised <- apply(EM_result_penguins$weights[na_indices, ], 1, function(x) {
  if (which.max(x) == 1) "Adelie" else "Gentoo"
})
```

```{r}
confusion_matrix_qda_supe = table(test_data$pred_supervised,test_data$species)
confusion_matrix_qda_supe
accuracy_qda_supe <- sum(diag(confusion_matrix_qda_supe)) / sum(confusion_matrix_qda_supe)
cat('Accuracy for Supervised GMM :- ',accuracy_qda_supe)
confusion_matrix_semi_supe = table(test_data$pred_semi_supervised,test_data$species)
confusion_matrix_semi_supe
accuracy_semi_supe <- sum(diag(confusion_matrix_semi_supe)) / sum(confusion_matrix_semi_supe)
cat('Accuracy for Semi Supervised GMM :- ',accuracy_semi_supe)
```

The difference in accuracy between the supervised and semi-supervised GMMM's can be attributed to the way each model handles labeled and unlabeled data.

1.  **Supervised GMM**: In the fully supervised setting, the model has access to complete labeled data, allowing it to directly estimate the parameters (means, covariances, and mixing proportions) for each class. This typically leads to more accurate parameter estimation and better classification performance since the model has direct access to the class labels.

2.  **Semi-Supervised GMM**: In the semi-supervised approach, only part of the data is labeled, while the rest is unlabeled. The semi-supervised GMM needs to iteratively estimate the missing labels and simultaneously update the parameters. This introduces uncertainty in the parameter estimation process, especially if the amount of labeled data is small. Misestimations in parameters can lead to suboptimal classification, as the model is less certain about the boundaries between classes.

In this case, the supervised GMM achieved higher accuracy with only one misclassification, while the semi-supervised GMM made four misclassifications. The semi-supervised model’s performance might improve with more labeled data, better initialization, or more sophisticated techniques for estimating the missing labels and variances.
